---
title: "Getting started with rstanTKTD"
author: "Virgile Baudrot and Sandrine Charles"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 4,
  cache = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```


```{r package, echo=FALSE, results='hide'}
library(rstanTKTD)
library(morse)
```

The package `rstanTKTD` is devoted to the analysis of data from standard toxicity
tests. It provides a simple workflow to calibrate GUTS models. This document illustrates
a typical use of `rstanTKTD` on survival data, which can be followed
step-by-step to analyze new datasets.

Analysis of results by visualization is possible with the package `morse`.

# TK-TD fitting with GUTS-SD, GUTS-IT, GUTS-PROPER lognormal and GUTS-PROPER loglogistic

## Loading data

Here is a typical session to analyse concentration-dependent time-course data
using the so-called GUTS models (Jager et al., 2011).

```{r data, cache=TRUE}
# (1) load dataset
data("data_Diazinon")

## ---- OPTIONAL - using package 'morse'
# (2) : check structure and integrity of the dataset, with the package morse
survDataCheck(data_Diazinon)

# (3) OPTIONAL - using package 'morse': represent the number of survivors as a function of time
plot(survData(data_Diazinon), pool.replicate = FALSE)

# (4) check information on the experimental design
summary(survData(data_Diazinon))
```

## Fitting models

- To fit the *Stochastic Death* model, GUTS-SD, we have to specify the `model_type` as `"SD"`.
- To fit the *Individual Tolerance* model, GUTS-IT, we have to specify the `model_type` as `"IT"`.
- To fit the general model, GUTS-PROPER, we have to specify the `model_type` as `"PROPER"` and the distribution `lognormal` or `loglogistic`.

Note that the implementation of the `IT`model is done with a loglogistic distribution.

```{r fit, cache=TRUE, echo=TRUE}
# # OPTION for the number of cores:
# options(mc.cores = 3)
#
# # (6) fit the TK-TD model SD
# fit_SD_diaz <- stan_guts(data_Diazinon, model_type = "SD")
# save(fit_SD_diaz, file = "tests/testdata/fit_SD_diaz.rda")
#
# # (7) fit the TK-TD model IT
# fit_IT_diaz <- stan_guts(data_Diazinon, model_type = "IT", warmup = 500, chains = 3)
# save(fit_IT_diaz, file = "tests/testdata/fit_IT_diaz.rda")
#
# # (8) fit the TK-TD model PROPER with distribution 'lognormal'
# fit_PROPERlogNormal_diaz <- stan_guts(data_Diazinon, model_type = "PROPER", distribution = "lognormal")
# save(fit_PROPERlogNormal_diaz, file = "tests/testdata/fit_PROPERlogNormal_diaz.rda")
#
# # (9) fit the TK-TD model PROPER with distribution 'loglogistic'
# fit_PROPERlogLogistic_diaz <- stan_guts(data_Diazinon, model_type = "PROPER", distribution = "loglogistic")
# save(fit_PROPERlogLogistic_diaz, file = "tests/testdata/fit_PROPERlogLogistic_diaz.rda")
load(file = "tests/testdata/fit_SD_diaz.rda")
load(file = "tests/testdata/fit_IT_diaz.rda")
load(file = "tests/testdata/fit_PROPERlogNormal_diaz.rda")
load(file = "tests/testdata/fit_PROPERlogLogistic_diaz.rda")
```

# Using internal functions for a quick plot

## Plot of survival rate

```{r rstanGUTS, cache = TRUE}
plot_stanguts(fit_SD_diaz)
plot_stanguts(fit_IT_diaz)
plot_stanguts(fit_PROPERlogNormal_diaz)
plot_stanguts(fit_PROPERlogLogistic_diaz)
```

## Plot of number of survivors

```{r rstanGUTS, cache = TRUE}
plot_stanguts(fit_SD_diaz, data_type = "Number")
plot_stanguts(fit_IT_diaz, data_type = "Number")
plot_stanguts(fit_PROPERlogNormal_diaz, data_type = "Number")
plot_stanguts(fit_PROPERlogLogistic_diaz, data_type = "Number")
```

# Using functions from package `rstan`

```{r stanfit, cache=TRUE}
stanfit_SD_diaz <- stanguts_to_stanfit(fit_SD_diaz)
stanfit_IT_diaz <- stanguts_to_stanfit(fit_IT_diaz)
stanfit_PROPERlogNormal_diaz <- stanguts_to_stanfit(fit_PROPERlogNormal_diaz)
stanfit_PROPERlogLogistic_diaz <- stanguts_to_stanfit(fit_PROPERlogLogistic_diaz)
```

## Print

```{r print, cache=TRUE}
print(stanfit_SD_diaz)
print(stanfit_IT_diaz)
print(stanfit_PROPERlogNormal_diaz)
print(stanfit_PROPERlogLogistic_diaz)
```

## Pairs

```{r pairs, cache=TRUE}
pairs(stanfit_SD_diaz, pars = c("hb_log10", "kd_log10", "z_log10", "kk_log10"))
pairs(stanfit_IT_diaz, pars = c("hb_log10", "kd_log10", "alpha_log10", "beta_log10"))
pairs(stanfit_PROPERlogNormal_diaz, pars = c("hb_log10", "kd_log10", "kk_log10", "alpha_log10", "beta_log10"))
pairs(stanfit_PROPERlogLogistic_diaz, pars = c("hb_log10", "kd_log10", "kk_log10", "alpha_log10", "beta_log10"))
```

# Using functions from package `morse (version >= 3.1.0)`

First of all, we have to convert the object of class `stanguts` into an object of class `survFit`:

```{r survFit, cache=TRUE}
survFit_SD_diaz <- stanguts_to_survFit(fit_SD_diaz)
survFit_IT_diaz <- stanguts_to_survFit(fit_IT_diaz)
survFit_PROPERlogNormal_diaz <- stanguts_to_survFit(fit_PROPERlogNormal_diaz)
survFit_PROPERlogLogistic_diaz <- stanguts_to_survFit(fit_PROPERlogLogistic_diaz)
```

## Summary

The `summary` function provides parameters estimates as medians and 95\% credible intervals.

```{r, cache=TRUE}
summary(survFit_SD_diaz)
summary(survFit_IT_diaz)
summary(survFit_PROPERlogNormal_diaz)
summary(survFit_PROPERlogLogistic_diaz)
```

## Plot

The `plot` function provides a representation of the fitting for each replicates

```{r, cache=TRUE}
plot(survFit_SD_diaz)
plot(survFit_IT_diaz)
plot(survFit_PROPERlogNormal_diaz)
plot(survFit_PROPERlogLogistic_diaz)
```

## PPC

The `ppc` function to check posterior prediction

```{r, cache=TRUE}
ppc(survFit_SD_diaz)
ppc(survFit_IT_diaz)
ppc(survFit_PROPERlogNormal_diaz)
ppc(survFit_PROPERlogLogistic_diaz)
```

Compared to the target time analysis, TK-TD modelling allows to compute and plot the lethal concentration for any *x* percentage and at any time-point. The chosen time-point can be specified with `time_LCx`, by default the maximal time-point in the dataset is used.

```{r cstSDLCx, cache=TRUE}
# LC50 at the maximum time-point:
LC50_SD_diaz <- LCx(survFit_SD_diaz, X = 50)
plot(LCx_SD_diaz)
# LC30 at the maximum time-4:
LC30t4_SD_diaz <- LCx(survFit_SD_diaz, X = 50, time_LCx = 4)
plot(LC30t4_SD_diaz)

# IT
LC50_IT_diaz <- LCx(survFit_IT_diaz, X = 50)
plot(LC50_IT_diaz)

# PROPER log-Normal
LC50_PROPERlogNormal_diaz <- LCx(survFit_PROPERlogNormal_diaz, X = 50)
plot(LC50_PROPERlogNormal_diaz)

# PROPER log-Logistic
LC50_PROPERlogLogistic_diaz <- LCx(survFit_PROPERlogLogistic_diaz, X = 50)
plot(LC50_PROPERlogLogistic_diaz)
```
