---
title: "Getting started with gutsRstan"
author: "Virgile Baudrot and Sandrine Charles"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
editor_options: 
  chunk_output_type: console
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 4,
  cache = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```


```{r package, echo=FALSE, results='hide'}
library(gutsRstan)
```

The package `gutsRstan` is devoted to the analysis of data from standard toxicity
tests. It provides a simple workflow to calibrate GUTS models (Jager et al., 2011 ; Jager and Ashauer, 2018). This document illustrates
a typical use of `gutsRstan` on survival data, which can be followed
step-by-step to analyze new data.

Analysis of results by visualization is possible with the package `morse`.

# Bayesian inference of GUTS models

There is only 4 functions in the `gutsRstan` package. We want to keep it simple.

* One function for the Bayesian inference of GUTS models. In this initial version we only impleted the GUTS-RED versions. We hope to implement  all other verions as exposed in the [GUTS book](https://leanpub.com/guts_book) in the future.
* One function to quickly plot the fit, compare your data with the fit.
* Two functions to transform a `stanguts` object into: (i) a `survFit` object to be used with the [`morse`](https://CRAN.R-project.org/package=morse) package ; or (ii) a `stanfit` object to use the set of functions of the [`rstan`](https://CRAN.R-project.org/package=rstan) package and all other functions from packages developped around the [`rstan`](https://CRAN.R-project.org/package=rstan) environment.

## Loading data

Here is a typical session to analyse concentration-dependent time-course data
using the General Unified Threshold model of Survival (GUTS) (see  Jager et al., 2011 or Jager and Ashauer, 2018).

```{r data, cache=TRUE}
# (1) load dataset
data("data_Diazinon")
```

The data set can be checked using the `morse` package where a function for GUTS model is also implemented using the JAGS language for Bayesian inference.

```{r morseChecking, cache=TRUE}
## ---- OPTIONAL - using 'morse' package 
library(morse)
# (2) : check structure and integrity of the dataset, with the package morse
survDataCheck(data_Diazinon)
# (3) OPTIONAL - using package 'morse': represent the number of survivors as a function of time
plot(survData(data_Diazinon), pool.replicate = FALSE)
# (4) check information on the experimental design
summary(survData(data_Diazinon))
```

## Fitting models

* To fit the *Stochastic Death* model, GUTS-SD, we have to specify the `model_type` as `"SD"`.
* To fit the *Individual Tolerance* model, GUTS-IT, we have to specify the `model_type` as `"IT"`.
* To fit the general model, GUTS-PROPER, we have to specify the `model_type` as `"PROPER"`.

Both model `IT` and `PROPER` may be fitted with either the `loglogistic` or the `lognormal` distribution. The default distribution is `loglogistic`. The `lognormal` distribution can be specified with the argument `distribution`.

Before fitting models, you can set the number of cores to use when executing the Markov chains in parallel, which defaults to 1 but we recommend setting the mc.cores option to be as many processors as
the hardware and RAM allow (up to the number of chains).

For instance, within `gutsRstan` the default number of chains is 3, so we recommend setting `mc.cores = 3`, if your machine has at least 3 cores which should be the case in a recent computer. 

```{r cores, cache=TRUE, echo=TRUE}
# (5) OPTION for the number of cores:
options(mc.cores = 3)
```

Fitting a data set is easy with the function `stan_guts()`.

The argument `model_type` is required with `"SD"`, `"IT"` or `"PROPER"`, there is no default value.

The argument `distribution` matters for `"IT"` or `"PROPER"` models. The default is `"loglogistic"`, the other distribution is `"lognormal"`. If you would like to test some other distributions not implemented yet, we would be happy to add any new ones: please let us know by opening a new [github issue](https://github.com/virgile-baudrot/gutsRstan/issues).

Here is the single line for fitting an `"SD"` model on data set `"data_Diazinon"`. We put all the outputs in an object named `fit_SD_diaz`.

```{r fit_SD, cache=TRUE, echo=TRUE}
# (6-SD) fit the TK-TD model SD
fit_SD_diaz <- stan_guts(data_Diazinon, model_type = "SD")
```

Here is the line for a fit of an `"IT"` model on data set `"data_Diazinon"`. The fit object is named `fit_IT_diaz`.

```{r fit_IT_ll, cache=TRUE, echo=TRUE, eval=FALSE}
# (6-IT_ll) fit the GUTS model IT (default: with distribution 'loglogistic')
fit_IT_diaz <- stan_guts(data_Diazinon, model_type = "IT")
```

The `IT` model can also be fitted with the `"lognormal"` distribution.

```{r fit_IT_ln, cache=TRUE, echo=TRUE, eval=FALSE}
# (6-IT_ln) fit the GUTS model IT with distribution 'lognormal'
fit_IT_lN_diaz <- stan_guts(data_Diazinon, model_type = "IT", distribution = "lognormal")
```

Here is the same line for a fit with thz `"PROPER"` model and the default `"loglogistic"` distribution.

```{r fit_PROPER_ll, cache=TRUE, echo=TRUE, eval=FALSE}
# (6-PROPER_ll) fit the GUTS model PROPER (default with distribution 'loglogistic')
fit_PROPER_ll_diaz <- stan_guts(data_Diazinon, model_type = "PROPER")
```

And finally, here is the line for a fit with `"PROPER"` model using the `"lognormal"` distribution.

```{r fit_PROPER_ln, cache=TRUE, echo=TRUE, eval=FALSE}
# (6-PROPER_ln) fit the GUTS model PROPER with distribution 'lognormal'
fit_PROPER_lN_diaz <- stan_guts(data_Diazinon, model_type = "PROPER", distribution = "lognormal")
```

# Plots of fitting results based on internal functions

## Plot of survival rate

```{r plotRate, cache = TRUE}
# Survival rate as a function of time
plot_stanguts(fit_SD_diaz)
```

```{r plotRate2, eval=FALSE}
plot_stanguts(fit_IT_diaz)
plot_stanguts(fit_IT_lN_diaz)
plot_stanguts(fit_PROPER_lN_diaz)
plot_stanguts(fit_PROPER_ll_diaz)
```

## Plot of number of survivors

```{r plotNumber, cache = TRUE}
# Number of survivors as a function of time
plot_stanguts(fit_SD_diaz, data_type = "Number")
```


```{r plotNumber2, cache = TRUE, eval=FALSE}
plot_stanguts(fit_IT_diaz, data_type = "Number")
plot_stanguts(fit_IT_lN_diaz, data_type = "Number")
plot_stanguts(fit_PROPER_lN_diaz, data_type = "Number")
plot_stanguts(fit_PROPER_ll_diaz, data_type = "Number")
```

# Plots of fitting results based on functions from the `rstan` package.

The function `stanguts_to_stanfit()` allows to use the output of the function `rstan_guts()` with packages developped around the R interface to the [Stan C++ library for Bayesian inference](http://mc-stan.org/)   [`rstan`](https://CRAN.R-project.org/package=rstan), [`shinystan`](https://CRAN.R-project.org/package=shinystan), [`bayesplot`](https://CRAN.R-project.org/package=bayesplot), [`loo`](https://CRAN.R-project.org/package=loo) packages 

```{r stanfit, cache=TRUE, eval=FALSE}
# Built of stanfit object
stanfit_SD_diaz <- stanguts_to_stanfit(fit_SD_diaz)
stanfit_IT_diaz <- stanguts_to_stanfit(fit_IT_diaz)
stanfit_IT_lN_diaz <- stanguts_to_stanfit(fit_IT_lN_diaz)
stanfit_PROPER_lN_diaz <- stanguts_to_stanfit(fit_PROPER_lN_diaz)
stanfit_PROPER_ll_diaz <- stanguts_to_stanfit(fit_PROPER_ll_diaz)
```

```{r loadRstan, echo=TRUE}
library(rstan)
```

## Explore the MCMC

From the R-package [`rstan`](https://CRAN.R-project.org/package=rstan), there are many functions to scrutinize the inference process. A simple one is the pairs plot:

```{r pairsRstan, cache=TRUE}
## Note: to reduce the size of the vignettes, we did not saved previous objects from the 'stanguts_to_stanfit()' function, so we recall this function in pairs plots.
# ---
pairs(stanguts_to_stanfit(fit_SD_diaz),
      pars = c("hb_log10", "kd_log10", "z_log10", "kk_log10"))
```

```{r pairsRstan2, eval=FALSE}
pairs(stanguts_to_stanfit(fit_IT_diaz),
      pars = c("hb_log10", "kd_log10", "alpha_log10", "beta_log10"))
pairs(stanguts_to_stanfit(fit_IT_lN_diaz),
      pars = c("hb_log10", "kd_log10", "alpha_log10", "beta_log10"))
pairs(stanguts_to_stanfit(fit_PROPER_lN_diaz),
      pars = c("hb_log10", "kd_log10", "kk_log10", "alpha_log10", "beta_log10"))
pairs(stanguts_to_stanfit(fit_PROPER_ll_diaz),
      pars = c("hb_log10", "kd_log10", "kk_log10", "alpha_log10", "beta_log10"))
```


# Explore outputs through Shinystan webpages

You can also explore the fitting results with the R-package [`shinystan`](https://CRAN.R-project.org/package=shinystan) :


```{r print, eval=FALSE}
library(shinystan)
launch_shinystan(stanfit_SD_diaz)
launch_shinystan(stanfit_IT_diaz)
launch_shinystan(stanfit_PROPER_lN_diaz)
launch_shinystan(stanfit_PROPER_ll_diaz)
```

# Plots of fitting results based on functions from package `morse`

At that time, `PROPER` models are not yet handled by the stable CRAN version of the [morse](https://CRAN.R-project.org/package=morse) package. So we only return output for `SD` and `IT` models. 

First of all, we have to convert the object of class `stanguts` into an object of class `survFit`:

```{r survFit, cache=TRUE, eval=FALSE}
survFit_SD_diaz <- stanguts_to_survFit(fit_SD_diaz)
survFit_IT_diaz <- stanguts_to_survFit(fit_IT_diaz)
```

## Summary

```{r loadmorse, echo=TRUE}
library(morse)
```

The `summary` function provides parameter estimates as medians and 95\% credible intervals.

```{r survFitsummary, cache=TRUE}
## Note: to reduce the size of the vignettes, we did not saved previous object from 'stanguts_to_survFit()' function, so we recall this function
# ---
summary(stanguts_to_survFit(fit_SD_diaz))
# summary(stanguts_to_survFit(fit_IT_diaz))
```

## Plot

The `plot` function provides a representation of the fitting results for each replicate

```{r survFitplot, cache=TRUE}
# plot(stanguts_to_survFit(fit_SD_diaz))
# plot(stanguts_to_survFit(fit_IT_diaz))
```

## PPC

The `ppc` function allows to check posterior predictions

```{r survFitppc, cache=TRUE}
ppc(stanguts_to_survFit(fit_SD_diaz))
# ppc(stanguts_to_survFit(fit_IT_diaz))
```

## Lethal concentration prediction 

Compared to the target time analysis, TKTD modelling allows to compute and plot the lethal concentration for any *x* percentage and at any time-point. The chosen time-point can be specified with `time_LCx` ; by default the maximal time-point in the data set is used.

```{r cstSDLCx, cache=TRUE}
# LC50 at the final time-point:
LC50_SD_diaz <- LCx(stanguts_to_survFit(fit_SD_diaz),
                    X = 50)
plot(LC50_SD_diaz)
# LC30 at the time 4:
LC30t4_SD_diaz <- LCx(stanguts_to_survFit(fit_SD_diaz),
                      X = 30, time_LCx = 4)
plot(LC30t4_SD_diaz)
```

## Multiplication Factor 

```{r profileMFx, cache=TRUE}
data_4predict <- data.frame(time = c(1, 2, 2.1, 5.9, 6, 7, 7.1, 10),
                           conc = c(10, 10, 0, 0, 10, 10, 0, 0))
plot(x = data_4predict$time, y = data_4predict$conc, type = "l",
     las = 1, xlab = "Time", ylab = "Concentration")
```


```{r cstSDMFx, cache=TRUE}
# MF50 at the maximum time-point:
MF50_SD_diaz <- MFx(stanguts_to_survFit(fit_SD_diaz),
                    data_predict = data_4predict, X = 50)
plot(MF50_SD_diaz)
# MF30 at the maximum time 20:
MF30t20_SD_diaz <- MFx(stanguts_to_survFit(fit_SD_diaz),
                      data_predict = data_4predict,
                      X = 30, time_MFx = 7)
plot(MF30t20_SD_diaz)
```

# References

Jager, T., Albert, C., Preuss, T. G. and Ashauer, R. (2011) *General unified threshold model of survival - a toxicokinetic-toxicodynamic framework for ecotoxicology*, Environmental Science & Technology, 45, 2529-2540.
 
Jager, T. and Ashauer, R. (2018) *Modelling survival under chemical stress. A comprehensive guide to the GUTS framework*. Version 1.0., *Leanpub*.